On exponential sums arising from the classical doubling method

### 2025-02-23 paper.tex
Calvin edits

\begin{theorem}\label{thm:computation-of-doubling-gauss-sum-scalar-for-deligne-lusztig-characters}
	Suppose that $T \cong \prod_{j=1}^r \multiplicativegroup{\finiteFieldExtension{k_j}} \times \prod_{i=1}^s \finiteFieldExtension{2m_i}^1 \left[\times \{\pm 1\}\right]$ is a maximal torus of $G$. Suppose that $\theta \colon T \to \multiplicativegroup{\cComplex}$ is a character and that under the above isomorphism $\theta = \alpha_1 \times \dots \times \alpha_r \times \theta_1 \times \dots \times \theta_s \times \epsilon$, where  $\left[\epsilon \colon \left\{\pm 1\right\} \to \multiplicativegroup{\cComplex},\right]$ $\alpha_j \colon \multiplicativegroup{\finiteFieldExtension{k_j}} \to \multiplicativegroup{\cComplex}$ and $\theta_i \colon \finiteFieldExtension{2m_i}^1 \to \multiplicativegroup{\cComplex}$ are characters for every $i$ and $j$. Denote $$g_T\left(\theta, \fieldCharacter\right) = \prod_{i=1}^s \GaussSumCharacter{\theta'_i}{\chi^{-1}}{\fieldCharacter_{2m_i}} \cdot  \prod_{j=1}^r \GaussSumCharacter{\alpha_j}{\chi^{-1}}{\fieldCharacter_{k_j}} \GaussSumCharacter{\alpha_j^{-1}}{\chi^{-1}}{\fieldCharacter_{k_j}}.$$ \calvin{we should add a $\chi$ as a variable to the left hand side of the above} Then
	\begin{align*}
		 \dblVirtualGaussSumScalar{R_{T, \theta}}{\chi} =& R_{T,\theta}\left(1\right) \GaussSumScalar{\chi^2}{\fieldCharacter}^{\left\lfloor\frac{\dim \hermitianSpace}{2}\right\rfloor} g_T\left(\theta, \fieldCharacter\right) \\
		& \times \prod_{i=1}^s \theta_i\left(-1\right) \cdot \prod_{j=1}^r \alpha_j\left(-1\right) \\
		& \left[\times \epsilon\left(-1\right) \chi\left(-2\right)\right],
	\end{align*}
	where the terms in $\left[\right]$ only appear in the odd orthogonal group case.
\end{theorem}
    \calvin{What does the $\chi(-2)$ come from? I didn't quite get that when I was messing around}
\begin{proof}
	This follows from the results mentioned above and from the formula $$R_T\left(1\right) = \varepsilon_{\algebraicGroup{G}} \varepsilon_{\algebraicGroup{T}} \frac{\grpIndex{G}{T}}{\sizeof{G}_p},$$ the fact that $$ \sizeof{\lieAlgebra} = \sizeof{G}_p^2 \cdot q^{\left\lfloor\frac{\dim \hermitianSpace}{2}\right\rfloor},$$
	the facts \begin{equation*}
		\varepsilon_{\algebraicGroup{T}} = \left(-1\right)^r \text{\,\,\,\,\,\, and \,\,\,\,\,\,}  \varepsilon_{\algebraicGroup{G}} = \left(-1\right)^{\left\lfloor\frac{\dim \hermitianSpace}{2}\right\rfloor},
	\end{equation*}
	and finally from $$\GaussSumScalar{\chi^2}{\fieldCharacter} = -\tau\left(\chi^2, \fieldCharacter\right).$$
\end{proof}
\begin{remark}
	Suppose that $\pi$ is any irreducible representation of $G$ such that $\trace \pi$ appears with non-zero coefficient in $R_{T,\theta}$. Then the central character of $\pi$ satisfies
	\begin{align*}
		\centralCharacter{\pi}\left(-1\right) = \theta\left(-1\right) =& \prod_{j=1}^r \alpha_j\left(-1\right) \cdot \prod_{i=1}^s \theta_i\left(-1\right)\\
		& \times \left[\epsilon\left(-1\right)\right].
	\end{align*}
	Thus \begin{align*}
		\dblVirtualGaussSumScalar{R_{T, \theta}}{\chi} =& R_{T,\theta}\left(1\right) \GaussSumScalar{\chi^2}{\fieldCharacter}^{\left\lfloor\frac{\dim \hermitianSpace}{2}\right\rfloor} \centralCharacter{\pi}\left(-1\right) g_T\left(\theta, \fieldCharacter\right) \\
		& \left[\times \chi\left(-2\right)\right].
	\end{align*}
\end{remark}
\subsection{Independence of geometric conjugacy for Gauss sums}
\calvin{here is the section I am adding in.}
\subsubsection{Geomtric Conjugacy and Lusztig Series}
(following chapter 13 of Digne and Michel's Representations of Finite Groups of Lie Type)
Let $\mathbb{G}$ be a reductive group over $\mathbb{F}$ and fix a torus $\mathbb{T}$ defined over $\mathbb{F}$. A \emph{dual group over $\mathbb{F}$} $\mathbb{G}^*$ with a dual torus $\mathbb{T}^*$ is a group with an isomorphism $X_*(T) \cong X^*(T^*)$ which is compatible with the action of Frobenius and sends roots to coroots \calvin{check if I flipped upper and lower star}.

Associated with a torus character pair $(T,\theta)$ is a  Frobenius stable $W(\mathbb{T})$ orbit in $X_*(\mathbb{T}) = X^*(\mathbb{T}^*)$. Under a choice of $\overline{\mathbb{F}}^\times \cong \left(\mathbb{Q}/\mathbb{Z}\right)_{p'}$, $X^*(\mathbb{T}^*) \otimes \left(\mathbb{Q}/\mathbb{Z}\right)_{p'} \cong \mathbb{T}^*(\overline{\mathbb{F}})$, this associates to $(T,\theta)$ a Frobenius stable semisimple conjugacy class in $\mathbb{G}^*$.

>>>>

\begin{theorem}\label{thm:computation-of-doubling-gauss-sum-scalar-for-deligne-lusztig-characters}
	Suppose that $T \cong \prod_{j=1}^r \multiplicativegroup{\finiteFieldExtension{k_j}} \times \prod_{i=1}^s \finiteFieldExtension{2m_i}^1 \left[\times \{\pm 1\}\right]$ is a maximal torus of $G$. Suppose that $\theta \colon T \to \multiplicativegroup{\cComplex}$ is a character and that under the above isomorphism $\theta = \alpha_1 \times \dots \times \alpha_r \times \theta_1 \times \dots \times \theta_s \times \epsilon$, where  $\left[\epsilon \colon \left\{\pm 1\right\} \to \multiplicativegroup{\cComplex},\right]$ $\alpha_j \colon \multiplicativegroup{\finiteFieldExtension{k_j}} \to \multiplicativegroup{\cComplex}$ and $\theta_i \colon \finiteFieldExtension{2m_i}^1 \to \multiplicativegroup{\cComplex}$ are characters for every $i$ and $j$. Denote $$g_T\left(\theta, \fieldCharacter\right) = \prod_{i=1}^s \GaussSumCharacter{\theta'_i}{\chi^{-1}}{\fieldCharacter_{2m_i}} \cdot  \prod_{j=1}^r \GaussSumCharacter{\alpha_j}{\chi^{-1}}{\fieldCharacter_{k_j}} \GaussSumCharacter{\alpha_j^{-1}}{\chi^{-1}}{\fieldCharacter_{k_j}}.$$ \calvin{we should add a $\chi$ as a variable to the left hand side of the above} Then
	\begin{align*}
		 \dblVirtualGaussSumScalar{R_{T, \theta}}{\chi} =& R_{T,\theta}\left(1\right) \GaussSumScalar{\chi^2}{\fieldCharacter}^{\left\lfloor\frac{\dim \hermitianSpace}{2}\right\rfloor} g_T\left(\theta, \fieldCharacter\right) \\
		& \times \prod_{i=1}^s \theta_i\left(-1\right) \cdot \prod_{j=1}^r \alpha_j\left(-1\right) \\
		& \left[\times \epsilon\left(-1\right) \chi\left(-2\right)\right],
	\end{align*}
	where the terms in $\left[\right]$ only appear in the odd orthogonal group case.
\end{theorem}
    \calvin{What does the $\chi(-2)$ come from? I didn't quite get that when I was messing around}
\begin{proof}
	This follows from the results mentioned above and from the formula $$R_T\left(1\right) = \varepsilon_{\algebraicGroup{G}} \varepsilon_{\algebraicGroup{T}} \frac{\grpIndex{G}{T}}{\sizeof{G}_p},$$ the fact that $$ \sizeof{\lieAlgebra} = \sizeof{G}_p^2 \cdot q^{\left\lfloor\frac{\dim \hermitianSpace}{2}\right\rfloor},$$
	the facts \begin{equation*}
		\varepsilon_{\algebraicGroup{T}} = \left(-1\right)^r \text{\,\,\,\,\,\, and \,\,\,\,\,\,}  \varepsilon_{\algebraicGroup{G}} = \left(-1\right)^{\left\lfloor\frac{\dim \hermitianSpace}{2}\right\rfloor},
	\end{equation*}
	and finally from $$\GaussSumScalar{\chi^2}{\fieldCharacter} = -\tau\left(\chi^2, \fieldCharacter\right).$$
\end{proof}
\begin{remark}
	Suppose that $\pi$ is any irreducible representation of $G$ such that $\trace \pi$ appears with non-zero coefficient in $R_{T,\theta}$. Then the central character of $\pi$ satisfies
	\begin{align*}
		\centralCharacter{\pi}\left(-1\right) = \theta\left(-1\right) =& \prod_{j=1}^r \alpha_j\left(-1\right) \cdot \prod_{i=1}^s \theta_i\left(-1\right)\\
		& \times \left[\epsilon\left(-1\right)\right].
	\end{align*}
	Thus \begin{align*}
		\dblVirtualGaussSumScalar{R_{T, \theta}}{\chi} =& R_{T,\theta}\left(1\right) \GaussSumScalar{\chi^2}{\fieldCharacter}^{\left\lfloor\frac{\dim \hermitianSpace}{2}\right\rfloor} \centralCharacter{\pi}\left(-1\right) g_T\left(\theta, \fieldCharacter\right) \\
		& \left[\times \chi\left(-2\right)\right].
	\end{align*}
\end{remark}
\subsection{Independence of geometric conjugacy for Gauss sums}
\calvin{here is the section I am adding in.}
\subsubsection{Geomtric Conjugacy and Lusztig Series}
(following chapter 13 of Digne and Michel's Representations of Finite Groups of Lie Type)
Let $\mathbb{G}$ be a reductive group over $\mathbb{F}$ and fix a torus $\mathbb{T}$ defined over $\mathbb{F}$. The \emph{dual group over $\mathbb{F}$} of $\mathbb{G}$ with respect to the torus $\mathbb{T}$ corresponds to the group $\mathbb{G}^*$ with a dual torus $\mathbb{T}^*$ and root datum
\[
    (X^*(\mathbb{T}^*),\Phi_{\mathbb{T}^*},X_*(\mathbb{T}^*),\Phi_{\mathbb{T}^*}^\vee) \cong (X_*(\mathbb{T}),\Phi_{\mathbb{T}}^\vee,X^*(\mathbb{T}),\Phi_{\mathbb{T}})
\]
such that the isomorphism respects the Frobenius action on $X^*$ and $X_*$.

Fix an isomorphism $\overline{\mathbb{F}}^\times \cong (\mathbb{Q}/\mathbb{Z})_{p'}$ and an embedding $\overline{\mathbb{F}}^\times \cong \overline{\mathbb{Q}_\ell}^\times$. By homing the exact sequence
\[
    0 \to T \to \mathbb{T}(\overline{\mathbb{F}}) \to_{F-1} \mathbb{T}(\overline{\mathbb{F}}) \to 0
\]
into $\mathbb{G}_m(\overline{\mathbb{F}})$, we get an exact sequence
\[
    1 \to X^*(\mathbb{T})_{(\mathbb{Q}/\mathbb{Z})_{p'}} \to_{F-1}X^*(\mathbb{T})_{(\mathbb{Q}/\mathbb{Z})_{p'}} \to \mathrm{Irr}(T) \to 1
\]
(the last map is surjective by including $T$ into $\mathbb{T}$ as a finite set of points and giving a surjection of group algebras).

Hence for a torus character pair $(T',\theta')$ for $G$, we may conjugate $\mathbb{T}'$ to a fixed torus $\mathbb{T}$ and associate to $\theta$ to an element of 
\[
\frac{(\mathbb{T})_{(\mathbb{Q}/\mathbb{Z})_{p'}}}{\{(\tilde{F}-1)x: x \in (\mathbb{T})_{(\mathbb{Q}/\mathbb{Z})_{p'}}\}}
\]
where $\tilde{F}$ is the Frobeinus twisted by a Weyl group element.

Frobenius stable $W(\mathbb{T})$ orbit in $X_*(\mathbb{T}) = X^*(\mathbb{T})$. Under a choice of $\overline{\mathbb{F}}^\times \cong \left(\mathbb{Q}/\mathbb{Z}\right)_{p'}$, $X^*(\mathbb{T}^*) \otimes \left(\mathbb{Q}/\mathbb{Z}\right)_{p'} \cong \mathbb{T}^*(\overline{\mathbb{F}})$, this associates to $(T,\theta)$ a Frobenius stable semisimple conjugacy class in $\mathbb{G}^*$.
